Analysing Constraint Grammars with a SAT solver

### 2015-10-23 paper/lrec/eval_futurework.tex
number of tag combinations

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
                      & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
Rules in grammar      & 59              & 279               & 1185     \\ \hline
Conflicts found       & 1               & 1                 & ???    \\ \hline
Time: all rules       & 7s              & 3min 22s          & 3h 26min    \\ \hline
Time: last rule       & 4s              & 7s                & 3min 40s    \\ \hline
\# variables: last rule & 33915         & 247153            &    \\ \hline
%Time: longest rule    & ?s              & ?s                & ?min     \\ \hline
\end{tabular}
\caption{Results}
\label{table:res}
\end{table}

\paragraph{Accuracy} We tested all three grammars if they contain non-applicable rules:
for each rule, apply all the preceding rules to the symbolic sentence, and then ask if that last rule can fire.
This test revealed problems in all grammars, and for the smaller grammars, we were able to verify manually that the results are true.

The Dutch grammar had the following rules in the given order, making the latter impossible to apply:

\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 Noun) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 Noun) ;
\end{verbatim}
\end{itemize}

The Spanish grammar had an erroneous set definition, where a word was required to have almost all POS tags at the same time. After fixing the set definition, our test didn't reveal other problems.
The program marked \todo{n} problem rules for the Finnish grammar, but we didn't have time to verify if all of them were genuinely problematic or if our program is at fault.
Correct errors include \todo{find a real problem and describe shortly}.
Our program detected tagset errors also in the Finnish grammar.
The grammar was originally written in 1995, and converted to the Omorfi tagset by \cite{pirinen2015}. 
For example, the Omorfi tagset marks a possessive suffix with \texttt{Px}, e.g. \texttt{PxPl2} for second person plural, but the grammar used the old convention \texttt{PL2} in the rules.
As a result, all rules with possessive suffixes were detected as impossible.
In addition, we suspect that some of the problems come just from our naive implementation.
For a condition that states $n$ or more words to the left or right, we only create a word in the sentence exactly $n$ words away. This is something to be fixed in the future.

 at least part of the result is related to the length of the sentences---for a condition that states $n$ or more words to the left or right, we only create a word exactly $n$ words away. This is something to be fixed in the future.

\paragraph{Running time} The speed of the check is an obvious issue currently.
The Finnish symbolic word has around 5700 variables in the beginning of the analysis, each representing a possible tag combination.
Around 3700 come from the Apertium morphological lexicon, excluding clitics, and around 2000 are present in the grammar---many of them include an actual word, both in target (\texttt{REMOVE (``bear'' verb) IF ...}) and in conditions (\texttt{REMOVE ... IF (-1 (``bear'' noun))}.

For Spanish and Dutch, the numbers are around 2000 and 300 respectively.
The Spanish lexicon includes up to two clitics (e.g. \emph{dámelo} `give.me.it') in the same word form, but Dutch has none in the lexicon.

>>>>

\begin{table}[]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
                      & \textsc{nld}  & \textsc{spa}  & \textsc{fin}  \\ \hline
\# rules in grammar      & 59              & 279      & 1185     \\ \hline
\# tag combinations      & 276             & 2191     & 5719    \\ \hline
\# conflicts found       & 1               & 1        & ???    \\ \hline
Time: all rules       & 7s              & 3min 22s    & 3h 26min    \\ \hline
Time: last rule       & 4s              & 7s          & 30min 25s    \\ \hline

% This is with just 2000ish variables for Finnish, above is with 5700
%Time: last rule       & 4s              & 7s                & 3min 40s    \\ \hline
%Time: longest rule    & ?s              & ?s                & ?min     \\ \hline
\# variables: last rule & 33915         & 247153            &    \\ \hline

\end{tabular}
\caption{Detecting rules that cannot be applied}
\label{table:res}
\end{table}

\paragraph{Correctness} We tested all three grammars if they contain non-applicable rules:
for each rule, apply all the preceding rules to the symbolic sentence, and then ask if that last rule can fire.
This test revealed problems in all grammars, and for the smaller grammars, we were able to verify manually that the results are true.

The Spanish grammar had an erroneous set definition, where a word was required to have almost all POS tags at the same time. After fixing the set definition, our test didn't reveal other problems.
The Dutch grammar had the following rules in the given order, making the latter impossible to apply:
\begin{itemize}
\item[] 
\begin{verbatim}REMOVE Adv IF (1 Noun) ;
REMOVE Adv IF (-1 Det) (0 Adj) (1 Noun) ;
\end{verbatim}
\end{itemize}

The program marked \todo{n} problem rules for the Finnish grammar, but we didn't have time to investigate all of them.
Correct errors include \todo{find a real problem and describe shortly}.
Our program detected also a number of tagset errors in the Finnish grammar.
The grammar was originally written in 1995, and converted to the Omorfi tagset by \cite{pirinen2015}. 
For example, the Omorfi tagset marks a possessive suffix with \texttt{Px}, e.g. \texttt{PxPl2} for second person plural, but the grammar used the old convention \texttt{PL2} in the rules.
As a result, all rules with possessive suffixes were detected as impossible.
In addition, we suspect that some of the problems come just from our naive implementation.
For a condition situated $n$ or more words to the left or right, we only create a word exactly $n$ words away. This is something to be fixed in the future.

\paragraph{Running time} The speed of the check is an obvious issue, especially for the Finnish grammar. The SAT problem grows the more rules

The Finnish symbolic word has around 5700 variables in the beginning of the analysis, each representing a possible tag combination.
Around 3700 come from the Apertium morphological lexicon, excluding clitics, and around 2000 are present in the grammar---many of them include an actual word, both in target (\texttt{REMOVE ("bear" verb) IF ...}) and in conditions (\texttt{REMOVE ... IF (-1 ("bear" noun))}.
For Spanish and Dutch, the numbers are around 2000 and 300 respectively.
The Spanish lexicon includes up to two clitics (e.g. \emph{dámelo} `give.me.it') in the same word form, but Dutch has none in the lexicon.

Not all questions are so expensive.
For instance, checking the tag sets in the Finnish grammar doesn't require the context of the other rules at all,
 The majority of the correctly detected errors in the Finnish grammar were rule-internal---these errors don't need the context of the other rules at all.
